---
title: "UKB KDSC CKD Staging Plots"
format: html
---

# 7. KDIGO CKD Staging Visualisation
Generate KDIGO heatmaps and Sankey diagrams from the CKD staging output.

## 0. Setup
```{r}
source(knitr::purl("R00-parameters.rmd", quiet = TRUE, output = tempfile()))

library(tidyverse)
library(data.table)
library(dtplyr)
library(arrow)
```

## 1. Load Data
```{r}
dt_timeline <- read_parquet("data/ckd_timeline.parquet") %>% as.data.table()
dt_cohort <- read_parquet("data/cohort.parquet") %>% as.data.table()
dt_visits <- read_parquet("data/visits.parquet") %>% as.data.table()

# Load baseline and 10y data
dt_baseline_all <- read_parquet("data/ckd_baseline.parquet") %>% as.data.table()
dt_10y <- read_parquet("data/ckd_10y.parquet") %>% as.data.table()

# Filter to GP-linked participants for transition analysis
dt_baseline <- dt_baseline_all %>%
  lazy_dt() %>%
  inner_join(dt_cohort %>% lazy_dt() %>% filter(has_gp) %>% select(eid), by = "eid") %>%
  as.data.table()
```

## 2. KDIGO Heatmaps
The KDIGO heatmap shows the distribution of CKD stages (G1-G5) by albuminuria
category (A1-A3). Cells are coloured by KDIGO risk: green (low), yellow
(moderate), orange (high), red (very high).

### 2.1 Baseline Heatmap
```{r fig.width=11.7, fig.height=8.3}
source("lib/plot_kdigo_grid.R")

baseline_plot <- plot_kdigo_heatmap(dt_baseline_all, "KDIGO Risk Classification at Baseline")
print(baseline_plot)

out_file <- "data/outputs/R07-kdigo_baseline.pdf"
dir.create(dirname(out_file), showWarnings = FALSE, recursive = TRUE)
pdf(out_file, width = 297 / 25.4, height = 210 / 25.4)
print(baseline_plot)
dev.off()
upload_to_dx(out_file, "kdsc_staging/outputs/R07-kdigo_baseline.pdf")
```

### 2.2 10-Year Heatmap (GP-linked)
```{r fig.width=11.7, fig.height=8.3}
tenyear_plot <- plot_kdigo_heatmap(dt_10y, "KDIGO Risk Classification at 10 Years Post-Baseline (GP-linked only)")
print(tenyear_plot)

out_file <- "data/outputs/R07-kdigo_10y.pdf"
pdf(out_file, width = 297 / 25.4, height = 210 / 25.4)
print(tenyear_plot)
dev.off()
upload_to_dx(out_file, "kdsc_staging/outputs/R07-kdigo_10y.pdf")
```

## 3. Sankey Diagrams: Stage Transitions (Baseline → 10y)
Track how CKD stages evolve from baseline to 10 years post-baseline among
GP-linked participants.

### 3.1 GFR Stage Transitions
```{r}
source("lib/plot_sankey.R")

sankey_stage <- plot_sankey_transitions(
  dt_baseline, dt_10y,
  column = "stage",
  start_label = "Baseline",
  end_label = "10y"
)

out_file <- "data/outputs/R07-sankey_stage.html"
dir.create(dirname(out_file), showWarnings = FALSE, recursive = TRUE)
save_sankey_html(sankey_stage, out_file)
# upload_to_dx(out_file, "kdsc_staging/outputs/R07-sankey_stage.html")

sankey_stage
```

### 3.2 KDIGO Stage Transitions
```{r}
sankey_kdigo <- plot_sankey_transitions(
  dt_baseline, dt_10y,
  column = "kdigo",
  start_label = "Baseline",
  end_label = "10y"
)

out_file <- "data/outputs/R07-sankey_kdigo.html"
save_sankey_html(sankey_kdigo, out_file)
# upload_to_dx(out_file, "kdsc_staging/outputs/R07-sankey_kdigo.html")

sankey_kdigo
```

### 3.3 KDIGO Risk Transitions
```{r}
sankey_risk <- plot_sankey_transitions(
  dt_baseline, dt_10y,
  column = "kdigo_risk",
  start_label = "Baseline",
  end_label = "10y"
)

out_file <- "data/outputs/R07-sankey_risk.html"
save_sankey_html(sankey_risk, out_file)
# upload_to_dx(out_file, "kdsc_staging/outputs/R07-sankey_risk.html")

sankey_risk
```

## 4. KDIGO Risk Distribution
```{r fig.width=8, fig.height=4}
kdigo_risk_colours <- c(
  "Low" = "#4CAF50",
  "Moderate" = "#FFC107",
  "High" = "#FF9800",
  "Very High" = "#F44336"
)
risk_levels <- c("Low", "Moderate", "High", "Very High")

risk_data <- dt_baseline %>%
  filter(kdigo_risk %in% risk_levels) %>%
  mutate(kdigo_risk = factor(kdigo_risk, levels = risk_levels)) %>%
  count(kdigo_risk)

ggplot(risk_data, aes(x = kdigo_risk, y = n, fill = kdigo_risk)) +
  geom_col() +
  scale_fill_manual(values = kdigo_risk_colours) +
  labs(
    title = "KDIGO Risk Distribution at Baseline",
    x = "KDIGO Risk Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 5. Stage Distribution Over Time
```{r fig.width=10, fig.height=5}
gfr_levels <- c("G1", "G2", "G3a", "G3b", "G4", "G5")

# Bin timeline dates into years relative to baseline
dt_by_year <- dt_timeline %>%
  lazy_dt() %>%
  left_join(
    dt_visits %>% lazy_dt() %>% filter(visit_index == 0) %>% select(eid, assessment_date),
    by = "eid"
  ) %>%
  filter(!is.na(stage)) %>%
  mutate(years_from_baseline = as.numeric(date - assessment_date) / 365.25) %>%
  mutate(year_bin = floor(years_from_baseline)) %>%
  filter(year_bin >= -5, year_bin <= 15) %>%
  as.data.table()

stage_by_year <- dt_by_year %>%
  group_by(eid, year_bin) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  count(year_bin, stage) %>%
  mutate(stage = factor(stage, levels = gfr_levels))

ggplot(stage_by_year, aes(x = year_bin, y = n, fill = stage)) +
  geom_area(position = "stack") +
  scale_fill_brewer(palette = "RdYlGn", direction = -1, name = "GFR Stage") +
  labs(
    title = "CKD Stage Distribution Over Time",
    x = "Years from Baseline",
    y = "Number of Participants"
  ) +
  theme_minimal()
```
